Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-08 tex/body.tex
tuning

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). The main insight is that a
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} \begin{longv}:
\fargtype{\tau} \to \tau\end{longv}$ converts the abstract type to the concrete
type. The translation preserves satisfiability and\begin{longv}, for admissible definitions,\end{longv}
unsatisfiability, and makes finite model finding possible for problems in this
class. Detailed proofs are included in a technical report \cite{our-report}.

@@

In addition to standard quantified formulas $\forall \vec x.\; \varphi$,
we consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
an uninterpreted function symbol. Their
semantics is the same as for standard quantified formulas.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1 : \tau_1$, $\ldots,$ $x_n : \tau_n$
and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
%I don't think it makes much sense to have any quantifiers within function bodies, 
%since they aren't operational, that's why I put quantifier-free. --ajr
We write $\exists
\vec x.\; \varphi$ as a shorthand for $\neg \forall \vec x.\; \neg \varphi$.

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$
can be shown well-founded under an strategy that
evaluates the condition of a $\lite$ before evaluating the relevant branch,
ignoring the other branch. Logically, such dependencies can be captured by
congruence rules. Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

\newcommand\badassex{%
 \{\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ),\;\,
 \forallf{\con{g}} x.\;\, \lite( \con{f}( x ) \teq 5,\;
   \con{g}( x ) \teq \con{f}( x ),\;
   \con{g}( x ) \teq \con{g}( x ) + 1)\}}

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern6.35em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
%\\[-1.25\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

@@

If $t$ is an application of a predicate $\boolop$, including the built-in
operators ${\neg}$, ${\wedge}$, ${\vee}$, ${\teq}$, and ${\lite}$,
$\conv_0$ calls itself recursively on the arguments $t_i$ and polarity $\pol(
\boolop, i, p )$, where $\pol$ is defined as
\[\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pnegate p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}\]
where the negation $\pnegate p$ of a polarity $p$ is $\pneg$ if $p$ is $\ppos$, $\ppos$
if $p$ is $\pneg$, and $\pnone$ if $p$ is $\pnone$.
If $t$ is positive, $\conv_0$ adds the constraint $\chi$ derived from
the subterms as a conjunction. Dually, if $t$ is negative, it constructs a
disjunction with the negated constraint. It $t$ has no polarity, the
constraint $\chi$ is simply returned to the caller.

If $t$ is a function definition of a function in
$\sfundefs{\Sigma}$, $\conv_0$ constructs a quantified formula over a single variable
$a$ of type $\fargtype{\con{f}}$ and replaces all occurrences of $\vec x$ in
$t'_1$ with $\vecfarg{\con{f}}( a )$. Since function definitions
are a top-level conjunct, %by case analysis on the return values of $\conv$,
$\chi$ must be $\ltrue$.
If $t$ is an unannotated quantified formula, $\conv_0$ calls itself
on the body with the same polarity. \begin{longv}A quantifier is prefixed to the
quantified formula and constraint returned by the recursive call.\end{longv}
Otherwise, $t$ is either an application of an uninterpreted predicate or a term
of a type other than $\Bool$. The returned constraint is a conjunction of
formulas of the form $\exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}(
a ) \teq \vec t$ for each subterm $\con{f}( \vec t )$ of $t$ such that $\con{f}
\in \sfundefs{\Sigma}$. Such constraints, when asserted positively, ensure that
the tuple of uninterpreted functions $\vecfarg{\con{f}}$ has $\vec t$
in its range.

@@

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, %we claim that
$\conv(\varphi)$ is equisatisfiable with $\varphi$ for any input $\varphi$.
%\ct{Must define "equisatisfiable".
%What we mean here is a weak form of satisfiability, we could instead say that
%$\conv(\varphi)$ is $T$-satisfiable iff $\varphi$ is $T$-satisfiable.
%} 
%\jb{We do define it, even it is weak form.}
The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable or unsatisfiable response from an SMT solver on input $\conv(
\varphi )$ implies the existence or nonexistence of a model for
$\varphi$. Moreover, the models of $\conv( \varphi )$ contain
pertinent information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) given above interprets $\con{c}$ as $14$
and $\con{s}(n) = \sum_{i=1}^n i$ for $0 \le n \le 14$,
and there exists a model of formula~(\ref{eq:ex-before}) that also interprets
$\con{c}$ and and $\con{s}(n)$ in the same way (for $0 \le n \le 14$).
In general, there exists a model of $\varphi$ that
coincides with each model of $\conv( \varphi )$ on its interpretation of all
functions in $\sfuns{\Sigma} - \sfundefs{\Sigma}$, i.e., $\{
\con{s} \}$ in the example. Furthermore, the model of $\conv( \varphi )$ will
also give correct information for the defined functions at all points belonging
to the domains of the corresponding abstract types. This can sometimes help
users debug their specifications.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. 
This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{\con{s}}$ can be
interpreted as the integers and $\farg{\con{s}}$ as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x.\;\, \varphi_\con{s} \}$ is admissible,
where $\varphi_\con{s}$ is the formula $\lite\bigl( x \leq 0,\; \con{s}(x) \teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
%As stated
%in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
%formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
%function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
%model.
%The rightmost conjunct $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$ 
%ensures that if $\con{s}( \con{c} ) > 100$ is a constraint that 
%formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
%satisfies $\varphi_{\con{s}}[\con{c}/x]$. 
%The nested existential
%$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
%a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
%any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. 
Clearly any interpretation $\I$ satisfying formula~(\ref{eq:ex-after}) satisfies 
$L_0 \cup \absconstraints( L_0 )$,
where $L_0 = \{ \con{s}( \con{c} ) > 100 \}$, 
where $\absconstraints( L_0 )$ is the set consisting of the single constraint $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$.
%The latter ensures that $\I$ satisfies a formula that entails both $\con{s}( \con{c} ) > 100$ and $\varphi_{\con{s}}[\con{c}/x]$.
Since $\I$ also satisfies both the translated function definition for $\con{s}$---i.e.,
the first conjunct of formula~(\ref{eq:ex-after})---and $\absconstraints( L_0
)$, it must also satisfy\begin{longv} the formula $\psi_1$\end{longv}:
\[
\lite\bigl( \con{c} \leq 0,\; 
            \con{s}(\con{c}) \teq 0,\;
            \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 )
            \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \con{c}-1 \bigr)
\]
The existential constraint in the above formula
ensures that whenever $\I$ satisfies the set $L_1 = L_0 \cup \{ \neg \con{c} \leq 0,\; \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 ) \}$,
$\I$ satisfies $\absconstraints( L_1 )$ as well.
Hence, by repeated application of this reasoning, it follows that
a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy% the formula
~$\psi$:
%
\[\con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!n-1} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq \con{c}-i + \con{s}( \con{c}-i-1 ) \bigr)
\wedge \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0\]
%
This formula is closed under function expansion
since it entails $\varphi_\con{s}[(\con{c}-i)/x]$ for $i = 0, \ldots, n$,
and it contains only $\con{s}$ applications corresponding to $\con{s}( \con{c}-i )$ for $i = 0, \ldots, n$.
Since $\{\forallf{\con{s}} x.\;\, \varphi_\con{s}\}$ is admissible,
there exists a $\Sigma$-interpretation satisfying $\psi \wedge \forallf{\con{s}} x.\;\, \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

Some finite model finders are based on a reduction to a decidable logic,
typically propositional logic. \begin{longv}They translate the input problem to the weaker
logic and pass it to a solver for that logic.\end{longv}
The translation is parameterized by upper or exact finite bounds on
the cardinalities of the atomic types. This procedure was pioneered by McCune
in the earlier versions of Mace (or MACE) \cite{mccune-1994}.
Other conceptually similar finders are Paradox \cite{claessen-sorensson-2003}
and FM-Darwin \cite{baumgartner-et-al-2009} for first-order logic with
equality; the Alloy Analyzer and its back-end Kodkod \cite{torlak-jackson-2007}
for first-order relational logic; and Refute \cite{weber-2008} and Nitpick
\cite{blanchette-nipkow-2010} for higher-order logic.

An alternative to translating the problem is to perform
an exhaustive model search directly on the original problem. Given fixed
cardinalities, the search space is represented in memory as multidimensional
arrays. The procedure tries different values in the function and predicate
table entries, checking each time if the problem is satisfied and backtracking
if necessary. This approach was pioneered by FINDER \cite{slaney-1994} and
SEM \cite{zhang-zhang-1995}\begin{longv} and serves as the basis of many more model finders,
notably the Alloy Analyzer's precursor \cite{jackson-1996} and the later
versions of~Mace \cite{mccune-prover9-mace4}\end{longv}.

@@

Model finding is concerned with satisfying arbitrary logical constraints. Some
tools are tailored for problems that correspond to total functional
programs. The QuickCheck tool for Haskell \cite{claessen-hughes-2000} is an
early example, based
on random testing. Bounded exhaustive testing \cite{runciman-et-al-2008} and
narrowing \cite{lindblad-2008-testing} are other successful strategies. These
tool are often much faster than model finders, but they typically cannot cope
with underspecification and nonexecutable functions.

>>>>

We define a translation for a class of
formulas involving admissible recursive function definitions
(Section~\ref{sec:encoding}). The main insight is that a
recursive definition $\forall x : \tau.\;\, \const{f}(x) \teq t$
can be translated to
$\forall a :
\fargtype{\tau}.\;\, \const{f}(\farg{\con{f}}(a)) \teq t[\concret(a)/x]$, where
$\fargtype{\tau}$ is an uninterpreted \relax{abstract} type and $\farg{\con{f}} \begin{longv}:
\fargtype{\tau} \to \tau\end{longv}$ converts the abstract type to the concrete
type. The translation preserves satisfiability and\begin{longv}, for admissible definitions,\end{longv}
unsatisfiability, and makes finite model finding possible for problems in this
class.\begin{shortv} Detailed proofs of correctness are included in a technical
report \cite{our-report}.\end{shortv}

@@

In addition to $\forall \vec x.\; \varphi$,
we consider \emph{annotated quantified formulas} of the form
$\forallf{\const{f}} \vec x.\; \varphi$, where $\con{f} \in \sfuns{\Sigma}$ is
uninterpreted. Their
semantics is the same as for standard quantified formulas.
Given $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables
$x_1 : \tau_1$, $\ldots,$ $x_n : \tau_n$
and $\varphi$ is a quantifier-free formula 
$T$-equivalent to $\con{f}( \vec x ) \teq t$ for some term $t$ of type $\tau$.
%\ct{Why can't $\varphi$ be existential instead of just quantifier free?} 
%I don't think it makes much sense to have any quantifiers within function bodies, 
%since they aren't operational, that's why I put quantifier-free. --ajr
%%\jb{Actually, in Isabelle we can have them and we use them sometimes. It certainly
%% makes sense. I can give you real examples if you like.}
We write $\exists
\vec x.\; \varphi$ as an abbreviation for $\neg \forall \vec x.\; \neg \varphi$.

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$
can be shown well-founded under an strategy that
evaluates the condition of a $\lite$ before evaluating the relevant branch,
ignoring the other branch. \begin{longv}Logically, such dependencies can be captured by
congruence rules. \end{longv}Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

\newcommand\badassex{%
 \{\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ),\;\,
 \forallf{\con{g}} x.\;\, \lite( \con{f}( x ) \teq 5,\allowbreak\;
   \con{g}( x ) \teq \con{f}( x ),\allowbreak\;
   \con{g}( x ) \teq \con{g}( x ) + 1)\}}

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern6.35em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
\\[-.8\baselineskip]
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
%\\[-1.25\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

@@

If $t$ is an application of a predicate $\boolop$, including the built-in
operators ${\neg}$, ${\wedge}$, ${\vee}$, ${\teq}$, and ${\lite}$,
$\conv_0$ calls itself recursively on the arguments $t_i$ and polarity $\pol(
\boolop, i, p )$, where $\pol$ is defined as
\[\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pnegate p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}\]
where the negation $\pnegate p$ of a polarity $p$ is $\pneg$ if $p$ is $\ppos$, $\ppos$
if $p$ is $\pneg$, and $\pnone$ if $p$ is $\pnone$.
If $t$ is positive, $\conv_0$ adds the constraint $\chi$ derived from
the subterms as a conjunction. Dually, if $t$ is negative, it constructs a
disjunction with the negated constraint. It $t$ has no polarity, the
constraint $\chi$ is simply returned to the caller.
%
\begin{longv}\par\end{longv}
%
If $t$ is a function definition of a function in
$\sfundefs{\Sigma}$, $\conv_0$ constructs a quantified formula over a single variable
$a$ of type $\fargtype{\con{f}}$ and replaces all occurrences of $\vec x$ in
$t'_1$ with $\vecfarg{\con{f}}( a )$. Since function definitions
are a top-level conjunct, %by case analysis on the return values of $\conv$,
$\chi$ must be $\ltrue$.
If $t$ is an unannotated quantified formula, $\conv_0$ calls itself
on the body with the same polarity. \begin{longv}A quantifier is prefixed to the
quantified formula and constraint returned by the recursive call.\end{longv}
Otherwise, $t$ is either an application of an uninterpreted predicate or a term
of a type other than $\Bool$. The returned constraint is a conjunction of
formulas of the form $\exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}(
a ) \teq \vec t$ for each subterm $\con{f}( \vec t )$ of $t$ such that $\con{f}
\in \sfundefs{\Sigma}$. Such constraints, when asserted positively, ensure that
the tuple of uninterpreted functions $\vecfarg{\con{f}}$ has $\vec t$
in its range.

@@

However, under the assumption that the function definitions in
$\sfundefs{\Sigma}$ are admissible, %we claim that
$\conv(\varphi)$ is equisatisfiable with $\varphi$ for any input $\varphi$.
%\ct{Must define "equisatisfiable".
%What we mean here is a weak form of satisfiability, we could instead say that
%$\conv(\varphi)$ is $T$-satisfiable iff $\varphi$ is $T$-satisfiable.
%} 
%\jb{We do define it, even it is weak form.}
The intuition is that the
interpretation of a term such as $\con{s}( -1 )$ in the previous example is not
relevant to the satisfiability of formula~(\ref{eq:ex-before}). Thus, a
satisfiable or unsatisfiable response from an SMT solver on input $\conv(
\varphi )$ implies the existence or nonexistence of a model for
$\varphi$. Moreover, the models of $\conv( \varphi )$ contain
pertinent information regarding the models of $\varphi$. For example, the model
$\I$ for formula~(\ref{eq:ex-after}) given above interprets $\con{c}$ as $14$
and $\con{s}(n)$ as $\sum_{i=1}^n i$ for $0 \le n \le 14$,
and there exists a model of formula~(\ref{eq:ex-before}) that also interprets
$\con{c}$ and $\con{s}(n)$ in the same way (for $0 \le n \le 14$).
In general, for every model of $\conv( \varphi )$,
there exists a model of $\varphi$ that
coincides with it on its interpretation of all
function symbols in $\sfuns{\Sigma} - \sfundefs{\Sigma}$ ($\{
\con{s} \}$ in the example). Furthermore, the model of $\conv( \varphi )$ will
also give correct information for the defined functions at all points belonging
to the domains of the corresponding abstract types. This can sometimes help
users debug their specifications.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. 
This formula is unsatisfiable only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{\con{s}}$ can be
interpreted as the integers and $\farg{\con{s}}$ as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable
only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x.\;\, \varphi_\con{s} \}$ is admissible,
where $\varphi_\con{s}$ is the formula $\lite\bigl( x \leq 0,\; \con{s}(x) \teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr)$.
%\ct{
%You cannot possibly mean "if and only if" in the two sentences above.
%Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
%(1) is sat iff (2) is sat. That is immediate.
%}
% You're right, changed both to "only if" --ajr
%As stated
%in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
%formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
%function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
%model.
%The rightmost conjunct $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$ 
%ensures that if $\con{s}( \con{c} ) > 100$ is a constraint that 
%formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
%satisfies $\varphi_{\con{s}}[\con{c}/x]$. 
%The nested existential
%$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
%a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
%any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. 
Clearly any interpretation $\I$ satisfying formula~(\ref{eq:ex-after}) satisfies 
$L_0 \cup \absconstraints( L_0 )$,
where $L_0 = \{ \con{s}( \con{c} ) > 100 \}$, 
where $\absconstraints( L_0 )$ is the set consisting of the single constraint $\exists a : {\fargtype{\con{s}}}. (\farg{\con{s}}( a ) \teq \con{c})$.
%The latter ensures that $\I$ satisfies a formula that entails both $\con{s}( \con{c} ) > 100$ and $\varphi_{\con{s}}[\con{c}/x]$.
Since $\I$ also satisfies both the translated function definition for $\con{s}$---i.e.,
the first conjunct of formula~(\ref{eq:ex-after})---and $\absconstraints( L_0
)$, it must also satisfy\begin{longv} the formula $\psi_1$\end{longv}
\[
\lite\bigl( \con{c} \leq 0,\; 
            \con{s}(\con{c}) \teq 0,\;
            \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 )
            \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \con{c}-1 \bigr)
\]
The existential constraint in the above formula
ensures that whenever $\I$ satisfies the set $L_1 = L_0 \cup \{ \neg \con{c} \leq 0,\; \con{s}(\con{c}) \teq \con{c} + \con{s}( \con{c}-1 ) \}$,
$\I$ satisfies $\absconstraints( L_1 )$ as well.
Hence, by repeated application of this reasoning, it follows that
a model of formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $n$ must also satisfy% the formula
~$\psi$:
%
\[\con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!n-1} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq \con{c}-i + \con{s}( \con{c}-i-1 ) \bigr)
\wedge \con{c}-n \leq 0 \wedge \con{s}( \con{c}-n ) \teq 0\]
%
This formula is closed under function expansion
since it entails $\varphi_\con{s}[(\con{c}-i)/x]$ for $i = 0, \ldots, n$,
and it contains only $\con{s}$ applications corresponding to $\con{s}( \con{c}-i )$ for $i = 0, \ldots, n$.
Since $\{\forallf{\con{s}} x.\;\, \varphi_\con{s}\}$ is admissible,
there exists a $\Sigma$-interpretation satisfying $\psi \wedge \forallf{\con{s}} x.\;\, \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

Some finite model finders are based on a reduction to a decidable logic,
typically propositional logic. \begin{longv}They translate the input problem to the weaker
logic and pass it to a solver for that logic.\end{longv}
The translation is parameterized by upper or exact finite bounds on
the cardinalities of the atomic types. This procedure was pioneered by McCune
in the earlier versions of
\begin{shortv}MACE\end{shortv}\begin{longv}Mace (or MACE)\end{longv}
\cite{mccune-1994}.
Other conceptually similar finders are Paradox \cite{claessen-sorensson-2003}
and FM-Darwin \cite{baumgartner-et-al-2009} for first-order logic with
equality; the Alloy Analyzer and its back-end Kodkod \cite{torlak-jackson-2007}
for first-order relational logic; and Refute \cite{weber-2008} and Nitpick
\cite{blanchette-nipkow-2010} for higher-order logic.

An alternative is to perform
an exhaustive model search directly on the original problem. Given fixed
cardinalities, the search space is represented as multidimensional
tables. The procedure tries different values in the function and predicate
tables, checking each time if the problem is satisfied.
This approach was pioneered by FINDER
\cite{slaney-1994} and SEM \cite{zhang-zhang-1995}\begin{longv} and serves as
the basis of many more model finders, notably the Alloy Analyzer's precursor
\cite{jackson-1996} and the later versions of~Mace
\cite{mccune-prover9-mace4}\end{longv}.

@@

Model finding is concerned with satisfying arbitrary logical constraints. Some
tools are tailored for problems that correspond to total functional
programs. The QuickCheck tool for Haskell \cite{claessen-hughes-2000} is an
early example, based on random testing. Bounded exhaustive
testing\begin{longv} \cite{runciman-et-al-2008}\end{longv} and
narrowing\begin{longv} \cite{lindblad-2008-testing}\end{longv} are other
successful strategies. These tool are often much faster than model finders, but
they typically cannot cope with underspecification and nonexecutable functions.
